---
title: "GDP components over time"
author: "Julien Vermeersch"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    highlight: zenburn
    number_sections: yes
    toc: yes
    toc_float: yes
    code_folding: show
---


```{r, setup, echo=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
options(digits = 3)

# default figure size
knitr::opts_chunk$set(
  fig.width=6.75, 
  fig.height=6.75,
  fig.align = "center"
)
```


```{r load-libraries, echo=FALSE}
library(tidyverse)  # Load ggplot2, dplyr, and all the other tidyverse packages
library(mosaic)
library(ggthemes)
library(GGally)
library(readxl)
library(here)
library(skimr)
library(janitor)
library(broom)
library(tidyquant)
library(infer)
library(openintro)
library(tidyquant)
library(kableExtra)
library(knitr)  
library(ggplot2)  
library(png) 
library(magick)
```

Welcome to my blog on the evolution of GDP and its components. In this section, I will discuss an analysis I did as part of my Data Analytics course at the London Business School. 

GDP can be divided into 4 main components, which are consumption (C), business investment (I), government spending (G) and net exports (exports - imports). You can read more about GDP and the different approaches in calculating at the [Wikipedia GDP page](https://en.wikipedia.org/wiki/Gross_domestic_product).

The GDP data I used is from the [United Nations' National Accounts Main Aggregates Database](https://unstats.un.org/unsd/snaama/Downloads), which contains estimates of total GDP and its components for all countries from 1970 to today. In this blog, I will look at how GDP and its components have changed over time, and compare different countries and how much each component contributes to that country's GDP. The file I will work with is [GDP and its breakdown at constant 2010 prices in US Dollars](http://unstats.un.org/unsd/amaapi/api/file/6).

```{r read_GDP_data}

UN_GDP_data  <-  read_excel(here::here("data", "Download-GDPconstant-USD-countries.xls"), # Excel filename
                sheet="Download-GDPconstant-USD-countr", # Sheet name
                skip=2) # Number of rows to skip

```

The first step of this analysis is to tidy the data. We want to change the data from its current wide format to a long, tidy format. 


```{r reshape_GDP_data}

tidy_GDP_data  <-  UN_GDP_data %>% 
  pivot_longer(
               cols="1970":"2017",
               names_to = "year", 
               values_to= "value"
               ) %>% 
  mutate(year= as.Date(year, "%Y"))

glimpse(tidy_GDP_data)

# Let us compare GDP components for these 3 countries
country_list <- c("United States","India", "Germany")
```

Once this has been done, we can start to split GDP up in its different components and examine how they have evolved over time in different countries. For this analysis, we will be looking at the United States, India and Germany. 

```{r, GDP_Components_Over_Time, fig1, fig.width=10, fig.height=6}

gdp_time <- tidy_GDP_data %>% 
  #filtering for the components of GDP and countries that we need
  filter(Country %in% country_list,
         IndicatorName %in% c("Gross capital formation", 
                              "Household consumption expenditure (including Non-profit institutions serving households)",
                              "Imports of goods and services",
                              "Exports of goods and services",
                              "General government final consumption expenditure")) %>% 
  mutate(value= value/ 1e9) 

gdp_time <- gdp_time  %>% 
  mutate(
    IndicatorName= case_when(
      IndicatorName == "Imports of goods and services" ~ "Imports",
      IndicatorName == "Exports of goods and services" ~ "Exports",
      IndicatorName == "General government final consumption expenditure" ~ "Government Expenditure",
      IndicatorName == "Household consumption expenditure (including Non-profit institutions serving households)" ~ "Household Expenditure",
      IndicatorName =="Gross capital formation"~"Gross Capital Formation"
    ))


ggplot(gdp_time, aes(x=year, y=value, color= IndicatorName)) +
  geom_line(aes(group=IndicatorName))+
  facet_wrap(~Country)+
  theme_bw()+
  labs(
    title = "GDP Components over time",
    subtitle= "In constant 2010 USD",
    y="Billion US$",
    color= "Components of GDP")+
  theme(axis.title.x=element_blank())


```

From these graphs we can see that different countries have had different GDP patterns over time. For example, in the United States imports increased at an incredibly rapid rate over the past couple of years. 

Next we calculate GDP from its basic components in order to asses the percentage difference between this figure and the GDP figure from the original dataframe. 

```{r, calculate GDP}

#calculating GDP as a sum of its components
GDP_Countries <- tidy_GDP_data %>% 
  #filtering for the components of GDP 
  filter(IndicatorName %in% c("Gross capital formation", 
                              "Household consumption expenditure (including Non-profit institutions serving households)",
                              "Imports of goods and services",
                              "Exports of goods and services",
                              "General government final consumption expenditure")) %>%  
  mutate(
    IndicatorName= case_when(
      IndicatorName == "Imports of goods and services" ~ "Imports",
      IndicatorName == "Exports of goods and services" ~ "Exports",
      IndicatorName == "General government final consumption expenditure" ~ "Government Expenditure",
      IndicatorName == "Household consumption expenditure (including Non-profit institutions serving households)" ~ "Household Expenditure",
      IndicatorName =="Gross capital formation"~"Gross Capital Formation"
    )) %>% 
pivot_wider(names_from= IndicatorName, values_from=value) %>% 
  #calculating Net Exports
  mutate(Net_Exports= Exports-Imports, 
         #extracting the year from the date 
         year=year(year)) %>% 
  select(-Imports, -Exports) %>% 
  pivot_longer(
    cols=4:7,
    names_to = "IndicatorName", 
    values_to= "value") %>% 
  group_by(year, Country) %>% 
  #summing the components of GDP and assigning to Gross_Domestic_Product
  summarise(Gross_Domestic_Product= sum(value))

GDP_Countries


```

```{r, % difference}
#selecting indicator GDP from UN GDP data
UN_GDP <- UN_GDP_data %>% 
  filter(IndicatorName=="Gross Domestic Product (GDP)") %>% 
  pivot_longer(
    cols="1970":"2017",
    names_to = "year", 
    values_to= "value"
  ) %>% 
  rename(United_Nations_Data_GDP= "value")

GDP_Countries <-
  GDP_Countries %>% 
  mutate(year=as.character(year)) 

#joining the UN GDP values with the calculated GDP and computing the percentage difference
GDP_diff<- inner_join(GDP_Countries, UN_GDP, by=c("Country", "year")) %>% 
  mutate(percentage_difference_in_GDP= (Gross_Domestic_Product-
                                          United_Nations_Data_GDP)/ United_Nations_Data_GDP) %>% 
  select(Country, year, percentage_difference_in_GDP)

paste("Percentage difference in GDP calculated vs GDP figure included in the dataframe:")

GDP_diff

```

The figures that result from our own calculations are pretty close to the actual figures, however there are some inor differences, which are shown in the table above. 


```{r, Proportion_GDP, fig2, fig.width=10, fig.height=6}


trans_UN_GDP_Data <-  UN_GDP_data %>% 
  #filtering for the components of GDP that we need for 3 selected countries
  filter(Country %in% country_list,
         IndicatorName %in% c("Gross capital formation", 
                              "Household consumption expenditure (including Non-profit institutions serving households)",
                              "Imports of goods and services",
                              "Exports of goods and services",
                              "General government final consumption expenditure",
                              "Gross Domestic Product (GDP)")) %>% 
  mutate(IndicatorName= case_when(
      IndicatorName == "Imports of goods and services" ~ "Imports",
      IndicatorName == "Exports of goods and services" ~ "Exports",
      IndicatorName == "General government final consumption expenditure" ~ "Government_Expenditure",
      IndicatorName == "Household consumption expenditure (including Non-profit institutions serving households)" ~ "Household_Expenditure",
      IndicatorName =="Gross capital formation"~"Gross_Capital_Formation",
      IndicatorName =="Gross Domestic Product (GDP)"~"GDP"
    )) %>% 
  pivot_longer(
    cols="1970":"2017",
    names_to = "year", 
    values_to= "value"
  ) %>% 
  mutate(year= as.Date(year, "%Y")) %>% 
  #changing to wide format to calculate the proportion of GDP for various components
  pivot_wider(names_from = IndicatorName, values_from= "value") %>% 
  mutate(
    Net_Exports= Exports- Imports,
    perc_NE= Net_Exports/ GDP,
    perc_GE= Government_Expenditure/ GDP,
    perc_GCF= Gross_Capital_Formation/GDP,
    perc_HE= Household_Expenditure/GDP
  ) %>% 
  #pivoting back to long format in order to create a time series plot
  pivot_longer(cols= "perc_NE": "perc_HE", names_to= "IndicatorName", values_to= "Proportion") %>% 
  mutate(
    IndicatorName= case_when(
      IndicatorName == "perc_HE" ~ "Household_Expenditure",
      IndicatorName =="perc_GCF"~"Gross_Capital_Formation",
      IndicatorName =="perc_NE"~"Net_Exports",
      IndicatorName== "perc_GE"~"Government_Expenditure"
    )) 

ggplot(trans_UN_GDP_Data, aes(x=year, y=Proportion, color= IndicatorName)) +
  geom_line(aes(group=IndicatorName))+
  facet_wrap(~Country)+
  theme_bw()+
  labs(
    title = "GDP and its breakdown at constant 2010 prices in US Dollars",
    y="proportion",
    color= "Components of GDP",
    caption="Source: United Nations , https://unstats.un.org/unsd/snaama/Downloads"
  )+
  theme(
    axis.title.x=element_blank(),
    legend.title = element_blank()
  )+
  scale_y_continuous(labels = scales::percent)+
  theme_bw()
      
      
```

This last chart compares the change in the proportion of the various components of GDP over time for three countries (United States, India and Germany). Since the early 2000s, Germany has increased the proportion of its GDP from Net Exports (Exports-Imports) and now has one of the highest trade surpluses in the world - this is an indicator of economic strength and competitiveness. 

In India proportion of GDP from Household Expenditure has decreased sharply while there has been a sharp increase in the proportion of its GDP from Gross Capital Formation - this shows that the economy is growing quickly and investing in fixed asset (roads, railways, machinery, etc.) 

Lastly, in the United States the proportion of GDP from Household Expenditure has increased while the Net Exports decreased, implying that US is slightly relying more on imports of goods. Also,  proportion of GDP from government expenditure has been slowly decreasing since 1970, indicating the reduced size of US government and its fiscal decentralization approach in the country.



