---
title: "GDP components over time"
author: "Julien Vermeersch"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    highlight: zenburn
    number_sections: yes
    toc: yes
    toc_float: yes
    code_folding: show
---


```{r, setup, echo=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
options(digits = 3)

# default figure size
knitr::opts_chunk$set(
  fig.width=6.75, 
  fig.height=6.75,
  fig.align = "center"
)
```


```{r load-libraries, echo=FALSE}
library(tidyverse)  # Load ggplot2, dplyr, and all the other tidyverse packages
library(mosaic)
library(ggthemes)
library(GGally)
library(readxl)
library(here)
library(skimr)
library(janitor)
library(broom)
library(tidyquant)
library(infer)
library(openintro)
library(tidyquant)
library(kableExtra)
library(knitr)  
library(ggplot2)  
library(png) 
library(magick)
```

# Challenge 2:GDP components over time and among countries

At the risk of oversimplifying things, the main components of gross domestic product, GDP are personal consumption (C), business investment (I), government spending (G) and net exports (exports - imports). You can read more about GDP and the different approaches in calculating at the [Wikipedia GDP page](https://en.wikipedia.org/wiki/Gross_domestic_product).

The GDP data we will look at is from the [United Nations' National Accounts Main Aggregates Database](https://unstats.un.org/unsd/snaama/Downloads), which contains estimates of total GDP and its components for all countries from 1970 to today. We will look at how GDP and its components have changed over time, and compare different countries and how much each component contributes to that country's GDP. The file we will work with is [GDP and its breakdown at constant 2010 prices in US Dollars](http://unstats.un.org/unsd/amaapi/api/file/6) and it has already been saved in the Data directory. Have a look at the Excel file to see how it is structured and organised


```{r read_GDP_data}

UN_GDP_data  <-  read_excel(here::here("data", "Download-GDPconstant-USD-countries.xls"), # Excel filename
                sheet="Download-GDPconstant-USD-countr", # Sheet name
                skip=2) # Number of rows to skip

```

The first thing you need to do is to tidy the data, as it is in wide format and you must make it into long, tidy format. Please express all figures in billions (divide values by `1e9`, or $10^9$), and you want to rename the indicators into something shorter.

> make sure you remove `eval=FALSE` from the next chunk of R code-- I have it there so I could knit the document

```{r reshape_GDP_data}

tidy_GDP_data  <-  UN_GDP_data %>% 
  pivot_longer(
               cols="1970":"2017",
               names_to = "year", 
               values_to= "value"
               ) %>% 
  mutate(year= as.Date(year, "%Y"))

glimpse(tidy_GDP_data)

# Let us compare GDP components for these 3 countries
country_list <- c("United States","India", "Germany")
```

GDP is the sum of Household Expenditure (Consumption *C*), Gross Capital Formation (business investment *I*), Government Expenditure (G) and Net Exports (exports - imports). We will calculate it given the components discussed above. 



```{r, GDP_Components_Over_Time, fig1, fig.width=10, fig.height=6}

gdp_time <- tidy_GDP_data %>% 
  #filtering for the components of GDP and countries that we need
  filter(Country %in% country_list,
         IndicatorName %in% c("Gross capital formation", 
                              "Household consumption expenditure (including Non-profit institutions serving households)",
                              "Imports of goods and services",
                              "Exports of goods and services",
                              "General government final consumption expenditure")) %>% 
  mutate(value= value/ 1e9) 

gdp_time <- gdp_time  %>% 
  mutate(
    IndicatorName= case_when(
      IndicatorName == "Imports of goods and services" ~ "Imports",
      IndicatorName == "Exports of goods and services" ~ "Exports",
      IndicatorName == "General government final consumption expenditure" ~ "Government Expenditure",
      IndicatorName == "Household consumption expenditure (including Non-profit institutions serving households)" ~ "Household Expenditure",
      IndicatorName =="Gross capital formation"~"Gross Capital Formation"
    ))


ggplot(gdp_time, aes(x=year, y=value, color= IndicatorName)) +
  geom_line(aes(group=IndicatorName))+
  facet_wrap(~Country)+
  theme_bw()+
  labs(
    title = "GDP Components over time",
    subtitle= "In constant 2010 USD",
    y="Billion US$",
    color= "Components of GDP")+
  theme(axis.title.x=element_blank())


```

```{r, calculate GDP}

#calculating GDP as a sum of its components
GDP_Countries <- tidy_GDP_data %>% 
  #filtering for the components of GDP 
  filter(IndicatorName %in% c("Gross capital formation", 
                              "Household consumption expenditure (including Non-profit institutions serving households)",
                              "Imports of goods and services",
                              "Exports of goods and services",
                              "General government final consumption expenditure")) %>%  
  mutate(
    IndicatorName= case_when(
      IndicatorName == "Imports of goods and services" ~ "Imports",
      IndicatorName == "Exports of goods and services" ~ "Exports",
      IndicatorName == "General government final consumption expenditure" ~ "Government Expenditure",
      IndicatorName == "Household consumption expenditure (including Non-profit institutions serving households)" ~ "Household Expenditure",
      IndicatorName =="Gross capital formation"~"Gross Capital Formation"
    )) %>% 
pivot_wider(names_from= IndicatorName, values_from=value) %>% 
  #calculating Net Exports
  mutate(Net_Exports= Exports-Imports, 
         #extracting the year from the date 
         year=year(year)) %>% 
  select(-Imports, -Exports) %>% 
  pivot_longer(
    cols=4:7,
    names_to = "IndicatorName", 
    values_to= "value") %>% 
  group_by(year, Country) %>% 
  #summing the components of GDP and assigning to Gross_Domestic_Product
  summarise(Gross_Domestic_Product= sum(value))

GDP_Countries


```

> What is the % difference between what you calculated as GDP and the GDP figure included in the dataframe?

```{r, % difference}
#selecting indicator GDP from UN GDP data
UN_GDP <- UN_GDP_data %>% 
  filter(IndicatorName=="Gross Domestic Product (GDP)") %>% 
  pivot_longer(
    cols="1970":"2017",
    names_to = "year", 
    values_to= "value"
  ) %>% 
  rename(United_Nations_Data_GDP= "value")

GDP_Countries <-
  GDP_Countries %>% 
  mutate(year=as.character(year)) 

#joining the UN GDP values with the calculated GDP and computing the percentage difference
GDP_diff<- inner_join(GDP_Countries, UN_GDP, by=c("Country", "year")) %>% 
  mutate(percentage_difference_in_GDP= (Gross_Domestic_Product-
                                          United_Nations_Data_GDP)/ United_Nations_Data_GDP) %>% 
  select(Country, year, percentage_difference_in_GDP)

paste("Percentage difference in GDP calculated vs GDP figure included in the dataframe:")

GDP_diff

```

> What is this last chart telling you? Can you explain in a couple of paragraphs the different dynamic among these three countries? 

This last chart compares the change in the proportion of the various components of GDP over time for three countries (United States, India and Germany). Since the early 2000s, Germany has increased the proportion of its GDP from Net Exports (Exports-Imports) and now has one of the highest trade surpluses in the world - this is an indicator of economic strength and competitiveness. 

In India proportion of GDP from Household Expenditure has decreased sharply while there has been a sharp increase in the proportion of its GDP from Gross Capital Formation - this shows that the economy is growing quickly and investing in fixed asset (roads, railways, machinery, etc.) 

Lastly, in the United States the proportion of GDP from Household Expenditure has increased while the Net Exports decreased, implying that US is slightly relying more on imports of goods. Also,  proportion of GDP from government expenditure has been slowly decreasing since 1970, indicating the reduced size of US government and its fiscal decentralization approach in the country. 


> Change list to include my own country: Belgium. 

```{r, Proportion_GDP, fig2, fig.width=10, fig.height=6}


trans_UN_GDP_Data <-  UN_GDP_data %>% 
  #filtering for the components of GDP that we need for 3 selected countries
  filter(Country %in% c("Italy", "Canada", "Belgium"),
         IndicatorName %in% c("Gross capital formation", 
                              "Household consumption expenditure (including Non-profit institutions serving households)",
                              "Imports of goods and services",
                              "Exports of goods and services",
                              "General government final consumption expenditure",
                              "Gross Domestic Product (GDP)")) %>% 
  mutate(IndicatorName= case_when(
      IndicatorName == "Imports of goods and services" ~ "Imports",
      IndicatorName == "Exports of goods and services" ~ "Exports",
      IndicatorName == "General government final consumption expenditure" ~ "Government_Expenditure",
      IndicatorName == "Household consumption expenditure (including Non-profit institutions serving households)" ~ "Household_Expenditure",
      IndicatorName =="Gross capital formation"~"Gross_Capital_Formation",
      IndicatorName =="Gross Domestic Product (GDP)"~"GDP"
    )) %>% 
  pivot_longer(
    cols="1970":"2017",
    names_to = "year", 
    values_to= "value"
  ) %>% 
  mutate(year= as.Date(year, "%Y")) %>% 
  #changing to wide format to calculate the proportion of GDP for various components
  pivot_wider(names_from = IndicatorName, values_from= "value") %>% 
  mutate(
    Net_Exports= Exports- Imports,
    perc_NE= Net_Exports/ GDP,
    perc_GE= Government_Expenditure/ GDP,
    perc_GCF= Gross_Capital_Formation/GDP,
    perc_HE= Household_Expenditure/GDP
  ) %>% 
  #pivoting back to long format in order to create a time series plot
  pivot_longer(cols= "perc_NE": "perc_HE", names_to= "IndicatorName", values_to= "Proportion") %>% 
  mutate(
    IndicatorName= case_when(
      IndicatorName == "perc_HE" ~ "Household_Expenditure",
      IndicatorName =="perc_GCF"~"Gross_Capital_Formation",
      IndicatorName =="perc_NE"~"Net_Exports",
      IndicatorName== "perc_GE"~"Government_Expenditure"
    )) 

ggplot(trans_UN_GDP_Data, aes(x=year, y=Proportion, color= IndicatorName)) +
  geom_line(aes(group=IndicatorName))+
  facet_wrap(~Country)+
  theme_bw()+
  labs(
    title = "GDP and its breakdown at constant 2010 prices in US Dollars",
    y="proportion",
    color= "Components of GDP",
    caption="Source: United Nations , https://unstats.un.org/unsd/snaama/Downloads"
  )+
  theme(
    axis.title.x=element_blank(),
    legend.title = element_blank()
  )+
  scale_y_continuous(labels = scales::percent)+
  theme_bw()
      
      
```





